\textbf{ Interpretable Gaussian Processes for Stellar Variability }

### 2021-01-12 tex/ms.tex
reorder yml

Another big advantage of GPs is their flexibility. GPs are often dubbed
a class of ``non-parametric'' models, given that nowhere in the specification
of the GP is there an explicit functional form for $\mathbbb{f}$. Rather, a GP
is a stochastic
process whose draws can in principle take on \emph{any} functional form,
subject, however, to certain smoothness and correlation criteria
of tunable strictness
that are fully
encoded in the covariance $\pmb{\Sigma}$.
%
In many applications, particularly when modeling stellar light curves,
it is customary to restrict the problem by
assuming that the process is \emph{stationary}, such that we may write
%
\begin{align}
    \label{eq:kernel}
    \Sigma_{i,j} & = k(t_i, t_j)
    \nonumber                                    \\
                 & = k(\left| t_i - t_j \right|)
    \nonumber                                    \\
                 & \equiv k(\Delta t)
    \quad.
\end{align}
%
A stationary process is one that is independent of phase (or, in this case,
the actual value of the time $t$); rather, it depends only on the \emph{difference}
between the phases of two data points. The kernel of a stationary process is
therefore a one-dimensional function, typically chosen from a set of
standard functions with desirable smoothness and spectral properties.
Among a vast number of kernels are a few that are commonly used
in modeling stellar variability: the exponential squared kernel,
%
\begin{align}
    k_\mathrm{E^2}(\Delta t)           & = \sigma^2 \exp\left(-\frac{\Delta t^2}{2\tau}\right)
    \quad,                                                                                                                                           \\
    \intertext{the Mat\'ern-3/2 kernel,}
    k_\mathrm{M^\frac{3}{2}}(\Delta t) & = \sigma^2 \left( 1 + \frac{\sqrt{3}\Delta t}{\tau} \right) \exp\left(-\frac{\sqrt{3}\Delta t}{\tau}\right)
    \quad,                                                                                                                                           \\
    \intertext{the exponential sine squared kernel,}
    k_\mathrm{S^2}(\Delta t)           & = \sigma^2 \exp\left[-\Gamma\sin^2\left( \frac{\pi \Delta t}{P} \right)\right]
    \quad,                                                                                                                                           \\
    \intertext{the cosine kernel,}
    k_\mathrm{C}(\Delta t)             & = \sigma^2 \cos\left( \frac{2\pi \Delta t}{P} \right)
    \quad,
\end{align}
%
or any linear combination and/or product of these. In the expressions above,
$\sigma$, $\tau$, $\Gamma$, and $P$ are
the \emph{hyperparameters} of the kernels.
Specifically, $\sigma^2$ is the variance (the value
of the kernel when $t_i = t_j$), $\tau$ is a typical timescale over which
data are correlated, $\Gamma$ is a dimensionless scale parameter, and $P$ is
a period. Once a kernel has been chosen, these scalar quantities fully
specify the covariance of the Gaussian process. GP-based inference
entails optimizing or sampling over these few quantities, which in many
cases can be done easily and efficiently.

@@

The two assumptions listed above suggest a fairly straightforward form for the
GP covariance in spherical harmonics and time:
%
\begin{align}
    \pmb{\Sigma}_\mathbf{t,y} = \pmb{\Sigma}_\mathbf{t} \otimes \pmb{\Sigma}_\mathbf{y}
    \quad,
\end{align}
%
where $\pmb{\Sigma}_\mathbf{t}$ is a $(K \times K)$ matrix describing the
covariance among the $K$ points in time,
$\pmb{\Sigma}_\mathbf{y}$ is the $(N \times N)$ matrix describing describing
the covariance among the $N \equiv (l_\mathrm{max} + 1)^2$ spherical
harmonic coefficients,
and $\otimes$ denotes the Kronecker product.
The quantity $\pmb{\Sigma}_\mathbf{t,y}$
is the $(NK \times NK)$ temporal-spatial covariance, whose coefficient at index
$(Nk + n, Nk' + n')$ is the covariance between the spherical harmonic coefficient
$y_{n}(t_k)$ and the spherical harmonic coefficient $y_{n'}(t_{k'})$.

\xxx{TODO: The model presented in this paper assumes statics stellar surfaces,
    corresponding to perfectly periodic light curves. This is hardly ever the
    case in practice, as spots evolve and differential rotation changes the surface
    over time. We can still use this model, but we need to either restrict our analysis
    to segments of light curves of duration shorter than the timescale on which
    the surface evolves, or explicitly model the effect of time evolution on the covariance.
    The next paper will deal with temporal evolution more
    rigorously, but the way to do this is just take the elementwise product of the \starryprocess
    flux covariance with a temporal covariance. This corresponds to the Kronecker product
    in spherical harmonic space. Show a figure, give an example, etc.}

@@

\begin{proof}{test_temporal}
    \pmb{\mathcal{A}}^\ddagger
    \left(\pmb{\Sigma}_\mathbf{t} \otimes \pmb{\Sigma}_\mathbf{y}\right)
    {\pmb{\mathcal{A}}^\ddagger}^\top
    &=
    \pmb{\Sigma} \odot \pmb{\Sigma}_\mathbf{t}
\end{proof}
%
where
%
\begin{align}
    \pmb{\mathcal{A}}^\ddagger
                      & \equiv
    \setstackgap{L}{1.25\baselineskip}
    \fixTABwidth{T}
    \parenMatrixstack{
    \mathbf{a}_0^\top &                   &        &                              \\
                      & \mathbf{a}_1^\top &        &                              \\
                      &                   & \ddots &                              \\
                      &                   &        & \,\,\, \mathbf{a}_{K-1}^\top
    }\quad.
\end{align}

@@

\subsection{Basic expression}
%
As we mentioned in \S\ref{sec:gp}, the flux $\mathbf{f}$ is a purely
linear function of the spherical harmonic coefficient vector $\mathbf{y}$:
%
\begin{align}
    \mathbf{f} = \mathbf{1} + \pmb{\mathcal{A}} \, \mathbf{y}
    \quad.
\end{align}
%
Even though this is derived in detail in \citet{Luger2019}, it is useful to
expand on the computation of the design matrix $\pmb{\mathcal{A}}$ in more
detail here. Let $\mathbf{a}_k^\top$ denote the $k^{th}$ row of $\pmb{\mathcal{A}}$,
such that
%
\begin{align}
    \label{eq:Arows}
    \pmb{\mathcal{A}}
     & =
    \setstackgap{L}{1.25\baselineskip}
    \fixTABwidth{T}
    \parenMatrixstack{
        \mathbf{a}_0^\top \\
        \mathbf{a}_1^\top \\
        \vdots            \\
        \mathbf{a}_{K-1}^\top
    }\quad.
\end{align}
%
The row vector $\mathbf{a}_k^\top$ encodes how the spherical harmonic
coefficient vector projects onto the $k^\mathrm{th}$ cadence in the flux timeseries, and
may be computed from
%
\begin{align}
    \label{eq:akT}
    \mathbf{a}_k^\top = \mathbf{r}^\top \,
    \mathbf{A_1} \,
    \mathbf{R}_{\hat{\mathbf{x}}}\left(-I\right) \,
    \mathbf{R}_{\hat{\mathbf{z}}}\left(\frac{2\pi}{P}t_k\right) \,
    \mathbf{R}_{\hat{\mathbf{x}}}\left(\frac{\pi}{2}\right)
    \quad.
\end{align}
%
To understand the expression above, let us proceed from right to left,
starting with the spherical harmonic vector $\mathbbb{y}$, which we assume
describes the surface intensity of the star at time $t = 0$
in a frame where $\hat{\mathbf{x}}$
points to the right, $\hat{\mathbf{y}}$ points up, and $\hat{\mathbf{z}}$
points out of the page. The quantity $\mathbf{R}_{\hat{\mathbf{x}}}$
is a Wigner rotation matrix
(described in detail in \S\ref{sec:wigner}),
which in this case rotates the spherical harmonic representation
of the star by an angle $\nicefrac{\pi}{2}$ counter-clockwise about $\hat{\mathbf{x}}$
such that the north pole of the star points along $\hat{\mathbf{z}}$. In this
frame, we apply a second Wigner rotation matrix, $\mathbf{R}_{\hat{\mathbf{z}}}$,
to rotate the star about $\hat{\mathbf{z}}$ counter-clockwise (i.e., eastward) by an angle
$\nicefrac{2\pi t_k}{P}$, where $P$ is the rotation period and $t_k$ is the
time at cadence $t$.
Next, we rotate the star by a \emph{clockwise} angle of $I$ about $\hat{\mathbf{x}}$,
where $I$ is the stellar inclination ($I = 0$ corresponding to a pole-on view and
$I = \nicefrac{\pi}{2}$ corresponding to an edge-on view). With this last rotation,
we are now in the observer's frame.%
\footnote{In principle, one last rotation could be performed about $\hat{\mathbf{z}}$
    to orient the projected disk of the star on the plane of the sky; however, the disk-integrated
    flux is independent of the rotation angle along the plane of the sky
    (which we refer to as the \emph{obliquity}), so this step is unnecessary.}

>>>>

Another big advantage of GPs is their flexibility. GPs are often dubbed
a class of ``non-parametric'' models, given that nowhere in the specification
of the GP is there an explicit functional form for $\mathbbb{f}$. Rather, a GP
is a stochastic
process whose draws can in principle take on \emph{any} functional form,
subject, however, to certain smoothness and correlation criteria
of tunable strictness
that are fully
encoded in the covariance $\pmb{\Sigma}$.
%
In many applications, particularly when modeling stellar light curves,
it is customary to restrict the problem by
assuming that the process is \emph{stationary}, such that we may write
%
\begin{align}
    \label{eq:kernel}
    \Sigma_{i,j} & = k(t_i, t_j)
    \nonumber                                    \\
                 & = k(\left| t_i - t_j \right|)
    \nonumber                                    \\
                 & \equiv k(\Delta t)
    \quad.
\end{align}
%
A stationary process is one that is independent of phase (or, in this case,
the actual value of the time $t$); rather, it depends only on the \emph{difference}
between the phases of two data points. The kernel of a stationary process is
therefore a one-dimensional function, typically chosen from a set of
standard functions with desirable smoothness and spectral properties.
Among a vast number of kernels are a few that are commonly used
in modeling stellar variability: the exponential squared kernel,
%
\begin{align}
    \label{eq:expsq}
    k_\mathrm{E^2}(\Delta t)           & = \sigma^2 \exp\left(-\frac{\Delta t^2}{2\tau}\right)
    \quad,                                                                                                                                           \\
    \intertext{the Mat\'ern-3/2 kernel,}
    \label{eq:mat32}
    k_\mathrm{M^\frac{3}{2}}(\Delta t) & = \sigma^2 \left( 1 + \frac{\sqrt{3}\Delta t}{\tau} \right) \exp\left(-\frac{\sqrt{3}\Delta t}{\tau}\right)
    \quad,                                                                                                                                           \\
    \intertext{the exponential sine squared kernel,}
    k_\mathrm{S^2}(\Delta t)           & = \sigma^2 \exp\left[-\Gamma\sin^2\left( \frac{\pi \Delta t}{P} \right)\right]
    \quad,                                                                                                                                           \\
    \intertext{the cosine kernel,}
    k_\mathrm{C}(\Delta t)             & = \sigma^2 \cos\left( \frac{2\pi \Delta t}{P} \right)
    \quad,
\end{align}
%
or any linear combination and/or product of these. In the expressions above,
$\sigma$, $\tau$, $\Gamma$, and $P$ are
the \emph{hyperparameters} of the kernels.
Specifically, $\sigma^2$ is the variance (the value
of the kernel when $t_i = t_j$), $\tau$ is a typical timescale over which
data are correlated, $\Gamma$ is a dimensionless scale parameter, and $P$ is
a period. Once a kernel has been chosen, these scalar quantities fully
specify the covariance of the Gaussian process. GP-based inference
entails optimizing or sampling over these few quantities, which in many
cases can be done easily and efficiently.

@@

The two assumptions listed above suggest a fairly straightforward form for the
GP covariance in spherical harmonics and time:
%
\begin{align}
    \label{eq:Sigmaty}
    \pmb{\Sigma}_\mathbf{y}^\mathbf{(t)} = \mathbf{K} \otimes \pmb{\Sigma}_\mathbf{y}
    \quad,
\end{align}
%
where $\mathbf{K}$ is a $(K \times K)$ matrix describing the
covariance among the $K$ points in time,
$\pmb{\Sigma}_\mathbf{y}$ is the $(N \times N)$ matrix describing describing
the covariance among the $N \equiv (l_\mathrm{max} + 1)^2$ spherical
harmonic coefficients (Equation~\ref{eq:cov_y}),
and $\otimes$ denotes the Kronecker product.
The quantity $\pmb{\Sigma}_\mathbf{y}^\mathbf{(t)}$
is the $(NK \times NK)$ temporal-spatial covariance, whose coefficient at index
$(Nk + n, Nk' + n')$ is the covariance between the spherical harmonic coefficient
$y_{n}(t_k)$ and the spherical harmonic coefficient $y_{n'}(t_{k'})$, where
the index $n$ is related to the spherical harmonic indices $l$ and $m$ via
Equation~(\ref{eq:n}). Finally, because of our assumption of stationarity,
the mean of the GP is still constant and equal to the mean of the static
process.

@@

The covariance matrix in Equation~(\ref{eq:Sigmaty}) can be sampled from to
yield time-variable surface maps, or it can be transformed into flux space
for sampling light curves or computing likelihoods. The covariance in
flux space is given by
%
\begin{align}
    \label{eq:SigmatSlow}
    \pmb{\Sigma}^\mathbf{(t)} & =
    \pmb{\mathcal{A}}^\ddagger
    \,
    \pmb{\Sigma}_\mathbf{y}^\mathbf{(t)}
    \,
    {\pmb{\mathcal{A}}^\ddagger}^\top
\end{align}
%
where
%
\begin{align}
    \pmb{\mathcal{A}}^\ddagger
                      & \equiv
    \setstackgap{L}{1.25\baselineskip}
    \fixTABwidth{T}
    \parenMatrixstack{
    \mathbf{a}_0^\top &                   &        &                              \\
                      & \mathbf{a}_1^\top &        &                              \\
                      &                   & \ddots &                              \\
                      &                   &        & \,\,\, \mathbf{a}_{K-1}^\top
    }
\end{align}
%
is a design matrix constructed by staggering the rows of the standard design
matrix $\pmb{\mathcal{A}}$ (see Appendix \ref{sec:starry:basic}). The diagonal
structure of $\pmb{\mathcal{A}}^\ddagger$ is due to the fact that each snapshot of the surface is only observed
at a single phase.

In the temporally-variable version of our GP, the covariance matrix in
Equation~(\ref{eq:SigmatSlow}) replaces the standard covariance
$\pmb{\Sigma}$; it can similarly be modified (\S\ref{sec:baseline})
to obtain the covariance of the normalized process. While $\pmb{\Sigma}^\mathbf{(t)}$
is $(K \times K)$, it is computed from the contraction of a much larger
$(NK \times NK)$ matrix, which is extremely inefficient to instantiate and
operate on. Fortunately, it can be shown that
%
\begin{proof}{test_temporal}
    \pmb{\Sigma}^\mathbf{(t)} & =
    \pmb{\mathcal{A}}^\ddagger
    \,
    \pmb{\Sigma}_\mathbf{y}^\mathbf{(t)}
    \,
    {\pmb{\mathcal{A}}^\ddagger}^\top
    \nonumber \\
    &=
    \pmb{\mathcal{A}}^\ddagger
    \left(\mathbf{K} \otimes \pmb{\Sigma}_\mathbf{y}\right)
    {\pmb{\mathcal{A}}^\ddagger}^\top
    \nonumber \\
    &=
    \pmb{\Sigma} \odot \mathbf{K}
    \quad,
\end{proof}
%
that is, the flux covariance is just the elementwise $\odot$ product of the
GP covariance $\pmb{\Sigma}$ and the temporal covariance $\mathbf{K}$.
This fact makes the temporal GP \emph{just as efficient} to
evaluate as the standard GP!

Armed with this algorithm for computing $\pmb{\Sigma}^\mathbf{(t)}$, it only
remains for us to decide on a structure for $\mathbf{K}$. While this can
in principle be any covariance matrix constructed from a stationary kernel,
we recommend one of the common radial kernels such as the exponential
squared kernel (Equation~\ref{eq:expsq}) or the Mat\'ern-3/2 kernel
(Equation~\ref{eq:mat32}) with variance parameter $\sigma^2$ set to
unity (since the variance is already specified within $\pmb{\Sigma}$). In this
case, the temporal covariance $\mathbf{K}$ is a function of a single
parameter: the timescale of the variability, $\tau$.

@@

\subsection{Basic expression}
%
\label{sec:starry:basic}
As we mentioned in \S\ref{sec:gp}, the flux $\mathbf{f}$ is a purely
linear function of the spherical harmonic coefficient vector $\mathbf{y}$:
%
\begin{align}
    \mathbf{f} = \mathbf{1} + \pmb{\mathcal{A}} \, \mathbf{y}
    \quad.
\end{align}
%
Even though this is derived in detail in \citet{Luger2019}, it is useful to
expand on the computation of the design matrix $\pmb{\mathcal{A}}$ in more
detail here. Let $\mathbf{a}_k^\top$ denote the $k^{th}$ row of $\pmb{\mathcal{A}}$,
such that
%
\begin{align}
    \label{eq:Arows}
    \pmb{\mathcal{A}}
     & =
    \setstackgap{L}{1.25\baselineskip}
    \fixTABwidth{T}
    \parenMatrixstack{
        \mathbf{a}_0^\top \\
        \mathbf{a}_1^\top \\
        \vdots            \\
        \mathbf{a}_{K-1}^\top
    }\quad.
\end{align}
%
The row vector $\mathbf{a}_k^\top$ encodes how the spherical harmonic
coefficient vector projects onto the $k^\mathrm{th}$ cadence in the flux timeseries, and
may be computed from
%
\begin{align}
    \label{eq:akT}
    \mathbf{a}_k^\top = \mathbf{r}^\top \,
    \mathbf{A_1} \,
    \mathbf{R}_{\hat{\mathbf{x}}}\left(-I\right) \,
    \mathbf{R}_{\hat{\mathbf{z}}}\left(\frac{2\pi}{P}t_k\right) \,
    \mathbf{R}_{\hat{\mathbf{x}}}\left(\frac{\pi}{2}\right)
    \quad.
\end{align}
%
To understand the expression above, let us proceed from right to left,
starting with the spherical harmonic vector $\mathbbb{y}$, which we assume
describes the surface intensity of the star at time $t = 0$
in a frame where $\hat{\mathbf{x}}$
points to the right, $\hat{\mathbf{y}}$ points up, and $\hat{\mathbf{z}}$
points out of the page. The quantity $\mathbf{R}_{\hat{\mathbf{x}}}$
is a Wigner rotation matrix
(described in detail in \S\ref{sec:wigner}),
which in this case rotates the spherical harmonic representation
of the star by an angle $\nicefrac{\pi}{2}$ counter-clockwise about $\hat{\mathbf{x}}$
such that the north pole of the star points along $\hat{\mathbf{z}}$. In this
frame, we apply a second Wigner rotation matrix, $\mathbf{R}_{\hat{\mathbf{z}}}$,
to rotate the star about $\hat{\mathbf{z}}$ counter-clockwise (i.e., eastward) by an angle
$\nicefrac{2\pi t_k}{P}$, where $P$ is the rotation period and $t_k$ is the
time at cadence $t$.
Next, we rotate the star by a \emph{clockwise} angle of $I$ about $\hat{\mathbf{x}}$,
where $I$ is the stellar inclination ($I = 0$ corresponding to a pole-on view and
$I = \nicefrac{\pi}{2}$ corresponding to an edge-on view). With this last rotation,
we are now in the observer's frame.%
\footnote{In principle, one last rotation could be performed about $\hat{\mathbf{z}}$
    to orient the projected disk of the star on the plane of the sky; however, the disk-integrated
    flux is independent of the rotation angle along the plane of the sky
    (which we refer to as the \emph{obliquity}), so this step is unnecessary.}
